---
title: "PBTA-ancestry demographic & clinical stats"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2023
params:
  plot_ci: TRUE
---

This script generates demographic and clinical summary statistics tables in the PBTA ancestry cohort 

```{r load packges}
library(data.table)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(colorblindr)
library(openxlsx)
```

Set directory and file paths

```{r set paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "demo-clin-stats")
results_dir <- file.path(analysis_dir, "results")

# set file paths
ancestry_file <- file.path(root_dir, "analyses", "survival", "results", "merged_ancestry_histology_survival.tsv")

source(file.path(analysis_dir, "util", "summary_functions.R"))
```

Wrangle data

```{r read ancestry}
hist <- read_tsv(ancestry_file) %>%
  mutate(age_at_diagnosis_days = as.numeric(age_at_diagnosis_days))
```

Consolidate unknown levels of factors, abbreviate long tumor histology

```{r}

# Define unknown levels of factors
unk_race <- c("Not Reported", "Other", "Reported Unknown", "Unknown")
unk_ethnicity <- c("Not Available", "Unavailable")
unk_resection <- c("Not Reported", "Unavailable")

hist <- hist %>%
    mutate(race = case_when(
      race %in% unk_race | is.na(race) ~ "Other/Unknown Race",
      grepl("White", race) ~ "White",
      TRUE ~ race
    )) %>%
    mutate(ethnicity = case_when(
      ethnicity %in% unk_ethnicity | is.na(ethnicity) ~ "Unknown",
      grepl("Not|Non", ethnicity) ~ "Not Hispanic or Latino",
      TRUE ~ ethnicity
    )) %>%
  mutate(reported_gender = case_when(
    is.na(reported_gender) | reported_gender %in% c("Not Reported", "Unknown") ~ "Unknown",
    TRUE ~ reported_gender
  )) %>%
  mutate(plot_group = case_when(
    plot_group == "Atypical Teratoid Rhabdoid Tumor" ~ "ATRT",
    TRUE ~ plot_group
  )) %>%
  mutate(extent_of_tumor_resection = case_when(
    extent_of_tumor_resection %in% unk_resection | is.na(extent_of_tumor_resection) ~ "Unavailable",
    grepl("Gross/Near", extent_of_tumor_resection) ~ "Gross/Near total resection",
    TRUE ~ extent_of_tumor_resection
  )) %>%
  mutate(cancer_predispositions = case_when(
    cancer_predispositions %in% c("None documented", "Not Reported") ~ "None documented",
    TRUE~ cancer_predispositions
  ))
```

Generate summary table of demographic and clinical features by predicted ancestry (full cohort)

```{r}
# Get count and percent of patients with and without P/LP variants
freq <- summarize_freq(hist, "predicted_ancestry")

# Look at distribution of following variables among patients with and without P/LP variants: sex, reported race, predicted ancestry, reported ethnicity, and histology group
gender <- summarize_count(hist, "predicted_ancestry", "reported_gender")

race <- summarize_count(hist, "predicted_ancestry", "race") %>%
  dplyr::select(-predicted_ancestry)

ethnicity <- summarize_count(hist, "predicted_ancestry", "ethnicity") %>%
  dplyr::select(-predicted_ancestry)

group <- summarize_count(hist, "predicted_ancestry", "plot_group") %>%
  dplyr::select(-predicted_ancestry)

resection <- summarize_count(hist, "predicted_ancestry", "extent_of_tumor_resection") %>%
  dplyr::select(-predicted_ancestry)

predisposition <- summarize_count(hist, "predicted_ancestry", "cancer_predispositions") %>%
  dplyr::select(-predicted_ancestry)

# merge stats and add %
merged <- gender %>%
  bind_cols(race, ethnicity, group, resection, predisposition) %>%
  column_to_rownames("predicted_ancestry") %>%
#  t() %>%
  as.data.frame()

perc <- round(merged / table(hist$predicted_ancestry) * 100, 1)
demo_path <- matrix(paste0(as.matrix(merged), " (", as.matrix(perc), "%)"),
               nrow = nrow(merged), ncol = ncol(merged),
               dimnames = list(rownames(merged), colnames(merged)))
demo_path <- ifelse(demo_path == "NA (NA%)", "--", demo_path)

demo_path <- demo_path %>%
  t() %>%
  as.data.frame()
  
# Calculate median survival and tmb for each group
clin <- hist %>%
  group_by(predicted_ancestry) %>%
  summarise(median_age = round(median(as.numeric(age_at_diagnosis_days)/365.25, na.rm = T), 2),
            median_age_at_chemo = round(median(age_at_chemo_start/365.25, na.rm = T), 2),
            median_age_at_radiation = round(median(age_at_radiation_start/365.25, na.rm = T), 2)) %>%
  column_to_rownames("predicted_ancestry") %>%
  t() %>%
  as.data.frame() %>%
  mutate_all(as.character)

# Merge stats and save
stats <- freq %>%
  bind_rows(demo_path, clin) %>%
  rownames_to_column("Term")

write_tsv(stats, 
          file.path(results_dir, "demo-clin-stats-all.tsv"))
```

Calculate corresponding Fisher's exact test p-values or ANOVA p-values for categorical and continuous variables, respectively. *NOTE: Fisher's exact test p-values are calculated using Monte Carlo simulation with N=10000 iterations due to tables sizes being larger than 2x2

```{r}
terms = c("Reported Gender", "Reported Race", "Reported Ethnicity", "Cancer Group",
          "Extent of Tumor Resection", "Cancer Predisposition", "Age at Diagnosis", 
          "Age at Chemo Start", "Age at Radiation Start")

gender_p <- fisher.test(table(hist$reported_gender, hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value

race_p <- fisher.test(table(hist$race, hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value

ethnicity_p <- fisher.test(table(hist$ethnicity, hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value

group_p <- fisher.test(table(hist$plot_group, hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value

resection_p <- fisher.test(table(hist$extent_of_tumor_resection, hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value

predisposition_p <- fisher.test(table(hist$cancer_predispositions, hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value

age_p <- anova(lm(age_at_diagnosis_days ~ predicted_ancestry, 
                  data = hist))$`Pr(>F)`[1]

chemo_p <- anova(lm(age_at_chemo_start/365.25 ~ predicted_ancestry, 
                  data = hist))$`Pr(>F)`[1]

radiation_p <- anova(lm(age_at_radiation_start/365.25 ~ predicted_ancestry, 
                  data = hist))$`Pr(>F)`[1]

pvals <- data.frame(term = terms,
                    pvalue = c(gender_p, race_p, ethnicity_p, 
                               group_p, resection_p, predisposition_p,
                               age_p, chemo_p, radiation_p))

write_tsv(pvals, file.path(results_dir, "demo-clin-pvals-all.tsv"))

```

Generate summary statistics tables within tumor histologies. Here we will include variables `mol_sub_group` and `CNS_region`, and exclude `plot_group`

```{r}

groups <- hist %>%
  filter(!is.na(plot_group) & !grepl("Other CNS embryonal tumor|Choroid plexus tumor", plot_group)) %>%
  pull(plot_group) %>%
  unique()

terms = c("Age at Diagnosis", "Reported Gender", "Reported Race", "Reported Ethnicity",
          "Molecular Subgroup", "CNS region", "Extent of Tumor Resection", 
          "Cancer predisposition syndrome", "Age at Chemo Start", "Age at Radiation Start")

stats_list <- list()

for (group in groups) {
  
  group_hist <- hist %>%
    filter(plot_group == group)

  # Get count and percent of patients with and without P/LP variants
  freq <- summarize_freq(group_hist, "predicted_ancestry")
  
  # Look at distribution of following variables among patients with and without P/LP variants: sex, reported race, predicted ancestry, reported ethnicity, and histology group
  gender <- summarize_count(group_hist, "predicted_ancestry", "reported_gender")
  
  race <- summarize_count(group_hist, "predicted_ancestry", "race") %>%
    dplyr::select(-predicted_ancestry)
  
  ethnicity <- summarize_count(group_hist, "predicted_ancestry", "ethnicity") %>%
    dplyr::select(-predicted_ancestry)
  
  subgroup <- summarize_count(group_hist, "predicted_ancestry", "mol_sub_group") %>%
    dplyr::select(-predicted_ancestry)
  
  region <- summarize_count(group_hist, "predicted_ancestry", "CNS_region") %>%
    dplyr::select(-predicted_ancestry)

  resection <- summarize_count(group_hist, "predicted_ancestry", "extent_of_tumor_resection") %>%
    dplyr::select(-predicted_ancestry)
  
  predisposition <- summarize_count(group_hist, "predicted_ancestry", "cancer_predispositions") %>%
    dplyr::select(-predicted_ancestry)
  
  # merge stats and add %
  merged <- gender %>%
    bind_cols(race, ethnicity, subgroup,
              region, resection, predisposition) %>%
    column_to_rownames("predicted_ancestry") %>%
  #  t() %>%
    as.data.frame()
  
  perc <- round(merged / table(group_hist$predicted_ancestry) * 100, 1)
  demo_path <- matrix(paste0(as.matrix(merged), " (", as.matrix(perc), "%)"),
                 nrow = nrow(merged), ncol = ncol(merged),
                 dimnames = list(rownames(merged), colnames(merged)))
  demo_path <- ifelse(demo_path == "NA (NA%)", "--", demo_path)
  
  demo_path <- demo_path %>%
    t() %>%
    as.data.frame()
    
  # Calculate median survival and tmb for each group
  clin <- group_hist %>%
    group_by(predicted_ancestry) %>%
    summarise(median_age = round(median(as.numeric(age_at_diagnosis_days)/365.25, na.rm = T), 2),
              median_age_chemo = round(median(age_at_chemo_start/365.25, na.rm = T), 2),
              median_age_radiation = round(median(age_at_radiation_start/365.25, na.rm = T), 2)) %>%
    column_to_rownames("predicted_ancestry") %>%
    t() %>%
    as.data.frame() %>%
    mutate_all(as.character)
  
  # Merge stats and save
  stats_list[[group]] <- freq %>%
    bind_rows(demo_path, clin) %>%
    rownames_to_column("Term")
  
}

# save df list to excel file
write.xlsx(stats_list,
           file.path(results_dir, "demo-clin-stats-by-histology.xlsx"),
           overwrite = TRUE,
           keepNA = TRUE,
           rowNames = FALSE)



```

Calculate corresponding p-values using same statistical test for each variable as was applied in tests with full cohort. 

```{r}

pvalue_list <- list()

for (group in groups) {
  
  group_hist <- hist %>%
    filter(plot_group == group)
  
  sex_p <- fisher.test(table(group_hist$reported_gender, group_hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value
  
  race_p <- fisher.test(table(group_hist$race, group_hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value
  
  ethnicity_p <- fisher.test(table(group_hist$ethnicity, group_hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value
  
  if (group %in% c("ATRT", "Low-grade glioma", "Mixed neuronal-glial tumor", "Ependymoma", "Medulloblastoma", "Other high-grade glioma")) {
    
    subgroup_p <- fisher.test(table(group_hist$mol_sub_group, group_hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value
    
  }
  
  region_p <- fisher.test(table(group_hist$CNS_region, group_hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value
  
  resection_p <- fisher.test(table(group_hist$extent_of_tumor_resection, group_hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value
  
  predisposition_p <- try(fisher.test(table(group_hist$cancer_predispositions, group_hist$predicted_ancestry),
                       simulate.p.value=TRUE, B = 10000)$p.value)
  predisposition_p <- ifelse(is.character(predisposition_p), NA, predisposition_p)
  
  age_p <- anova(lm(age_at_diagnosis_days ~ predicted_ancestry, 
                    data = group_hist))$`Pr(>F)`[1]
  
  chemo_p <- try(anova(lm(age_at_chemo_start/365.25 ~ predicted_ancestry, 
                    data = group_hist))$`Pr(>F)`[1])
  chemo_p <- ifelse(is.character(chemo_p), NA, chemo_p)
  
  radiation_p <- try(anova(lm(age_at_radiation_start/365.25 ~ predicted_ancestry, 
                    data = group_hist))$`Pr(>F)`[1])
  radiation_p <- ifelse(is.character(radiation_p), NA, radiation_p)
    
  
  if (group %in% c("ATRT", "Low-grade glioma", "Mixed neuronal-glial tumor", "Ependymoma", "Medulloblastoma", "Other high-grade glioma")) {
    
    pvalue_list[[group]] <- data.frame(term = terms,
                                    pvalue = c(age_p, sex_p, race_p, 
                                               ethnicity_p, subgroup_p, 
                                               region_p, resection_p,
                                               predisposition_p,
                                               chemo_p, radiation_p))
    
  } else {
  
    pvalue_list[[group]] <- data.frame(term = terms[terms != "Molecular Subgroup"],
                                    pvalue = c(age_p, sex_p, race_p, 
                                               ethnicity_p,
                                               region_p, resection_p,
                                               predisposition_p,
                                               chemo_p, radiation_p))
  }
  
}

write.xlsx(pvalue_list,
           file.path(results_dir, "demo-clin-pvalues-by-histology.xlsx"),
           overwrite = TRUE,
           keepNA = TRUE,
           rowNames = TRUE)
                    
```

Print session info
```{r session info}
sessionInfo()

```