---
title: "DEI Survival Analyses"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2023
---

**Purpose:** 

Runs survival analysis models within tumor histologies and molecular subtypes by predicted ancestry, self-reported race, and self-reported ethnicity

## Setup

#### Packages and functions

Read in set up script.

```{r Set up library}
library(survival)
library(ggpubr)
library(tidyverse)

# Set directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "survival")
# Source this script, which contains a wrapper function that can conduct the survival analyses, from OpenPBTA
source(file.path(analysis_dir, "util", "survival_models.R"))
# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

#### Set up files and directories

Set up output directories. 
```{r Set up directories}
input_dir <- file.path(root_dir, "analyses", "survival", "results")
results_dir <- file.path(analysis_dir, "results")
plots_dir <- file.path(analysis_dir, "plots")
```

Make output directories.
```{r Make output directories}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}
```

Declare input file paths and useful variables
```{r Set input file paths}
ancestry_file <- file.path(input_dir, "merged_ancestry_histology_survival.tsv")
```


## Import the metadata
```{r Read in metadata}
metadata <- readr::read_tsv(ancestry_file)

metadata <- metadata %>%
    mutate(mol_sub_group = case_when(
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("wildtype", molecular_subtype) ~ "HGG, H3 WT",
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("K28", molecular_subtype) ~ "DMG, H3 K28",
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("G35", molecular_subtype) ~ "DHG, H3 G35",
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("IHG", molecular_subtype) ~ "IHG",
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("IDH", molecular_subtype) ~ "HGG, IDH",
    plot_group %in% c("Low-grade glioma") & grepl("SEGA", molecular_subtype) ~ "SEGA",
    plot_group %in% c("Low-grade glioma") & grepl("-BRAF", molecular_subtype) ~ "LGG, BRAF fusion",
    plot_group %in% c("Low-grade glioma") & grepl("wildtype", molecular_subtype) ~ "LGG, WT",
    plot_group %in% c("Low-grade glioma") & grepl("To be classified", molecular_subtype) ~ "LGG, To be classified",
    plot_group %in% c("Low-grade glioma") ~ "LGG, Other alteration",
    plot_group %in% c("Mixed neuronal-glial tumor") & grepl("wildtype", molecular_subtype) ~ "GNG/GNT, WT",
    plot_group %in% c("Mixed neuronal-glial tumor") & grepl("-BRAF", molecular_subtype) ~ "GNG/GNT, BRAF fusion",
    plot_group %in% c("Mixed neuronal-glial tumor") & !grepl("To be classified", molecular_subtype) ~ "GNG/GNT, Other alteration",
    !is.na(molecular_subtype) ~ molecular_subtype
  )) %>%
  mutate(age_at_diagnosis_days = as.numeric(age_at_diagnosis_days)) %>%
  mutate(age_at_diagnosis_years = age_at_diagnosis_days/365.25)
```

Modify reported race and ethnicity to only include 
```{r}
metadata <- metadata %>%
  dplyr::mutate(race = case_when(
    grepl("White", race) ~ "White",
    grepl("Black", race) ~ "Black/Afr American",
    grepl("Asian", race) ~ "Asian",
    TRUE ~ "Other/Unknown"
  )) %>%
  dplyr::mutate(race = fct_relevel(race,
                                   c("Black/Afr American",
                                     "White", "Asian", 
                                     "Other/Unknown"))) %>%
  dplyr::mutate(ethnicity = case_when(
    grepl("Non-Hispanic|Not Hispanic", ethnicity) ~ "Not Hispanic/Latino",
    grepl("Hispanic", ethnicity) ~ "Hispanic/Latino",
    TRUE ~ NA_character_
  )) %>%
  dplyr::mutate(ethnicity = fct_relevel(ethnicity,
                                        c("Not Hispanic/Latino",
                                          "Hispanic/Latino")))

variable <- c("predicted_ancestry", "race", "ethnicity", "EUR_status")
```

Define plot groups for which survival analyses should be run
```{r}
# Define groups for survival analyses
groups <- c("Atypical Teratoid Rhabdoid Tumor", "Craniopharyngioma",
            "DIPG or DMG|Other high-grade glioma", "Ependymoma",
            "Mixed neuronal-glial tumor", "Low-grade glioma",
            "Medulloblastoma", "Meningioma",
            "Mesenchymal tumor", "Neurofibroma plexiform", 
            "Oligodendroglioma", "Schwannoma"
)

# create names vector for output files
plot_names <- c("atrt", "cranio", "hgg", "epn", "gng", "lgg", "mb", "mng", "mes", "nfp", "oligo", "swn")
names(plot_names) <- groups
```

Loop through groups and create Kaplan Meier survival and Cox Proportional Hazard model for OS and EFS
```{r}

for (var in variable){

  for (group in groups){
    
    results_dir <- file.path(analysis_dir, "results", plot_names[group])
    
    if (!dir.exists(results_dir)) {
      dir.create(results_dir)
      
      }
    
    km_os_model_file <- file.path(results_dir, 
                                      glue::glue("logrank_{plot_names[group]}_OS_{var}.RDS"))

    coxph_os_model_file <- ifelse(group %in% c("Low-grade glioma"),
                               file.path(results_dir,
                                         glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_resection_{var}.RDS")),
                               file.path(results_dir,
                                         glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_{var}.RDS"))
    )
    
    km_efs_model_file <- file.path(results_dir, 
                                      glue::glue("logrank_{plot_names[group]}_EFS_{var}.RDS"))

    coxph_efs_model_file <- ifelse(group %in% c("Low-grade glioma"),
                               file.path(results_dir,
                                         glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_resection_{var}.RDS")),
                               file.path(results_dir,
                                         glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_{var}.RDS"))
    )
    
    subtype_plot_pdf <- file.path(plots_dir, 
                                  glue::glue("{plot_names[group]}_subtype_by_{var}.pdf"))
    
    # Subset histology
    group_hist <- metadata %>%
      filter(grepl(group, plot_group))
    
    if (group == "DIPG or DMG|Other high-grade glioma"){
      group_hist <- group_hist %>%
       mutate(mol_sub_group = fct_relevel(mol_sub_group, c("HGG, H3 WT",
                                                       "DHG, H3 G35", "DMG, H3 K28",
                                                       "IHG", "HGG, IDH")))
    } 
    
    if (group == "Ependymoma") {
      group_hist <- group_hist %>%
        dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                      c("EPN, ST ZFTA", "EPN, ST YAP1", "EPN, PF A", 
                        "EPN, PF B", "EPN, MPE", "EPN, SP", "EPN, SP-MYCN",
                        "EPN, To be classified")))
        
    } 
    
    if (group == "Low-grade glioma") {
      group_hist <- group_hist %>%
        dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                                                  c("LGG, WT", "LGG, BRAF fusion",
                                                    "LGG, Other alteration",
                                                    "SEGA",
                                                    "LGG, To be classified")))
    }
    
    if (group == "Medulloblastoma") {
      group_hist <- group_hist %>%
        dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                                                  c("MB, WNT", "MB, Group3",
                                                    "MB, Group 4", "MB, SHH")))
    }
    
    if (group == "Mixed neuronal-glial tumor") {
      group_hist <- group_hist %>%
        dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                                                  c("GNG/GNT, WT", "GNG/GNT, BRAF fusion",
                                                    "GNG/GNT, Other alteration",
                                                    "GNG, To be classified",
                                                    "GNT, To be classified")))
    }
    
    if (group %in% c("DIPG or DMG|Other high-grade glioma",
                     "Mixed neuronal-glial tumor", "Low-grade glioma")) {

      subtype_plot <- group_hist %>%
              filter(!is.na(mol_sub_group)) %>%
              dplyr::count(predicted_ancestry, mol_sub_group, name = "count", .drop = FALSE) %>%
              ggplot(aes(x = predicted_ancestry, y = mol_sub_group, fill = count)) +
              geom_tile(color = "black",
                        lwd = 1,
                        linetype = 1, show.legend = FALSE) +
              scale_fill_gradient(low="white", high="orangered") +
              geom_text(aes(label = count), color = "black", size = 4) +
              labs(x = NULL, y = NULL) +
              coord_fixed() +
              theme_minimal()
      ggsave(subtype_plot_pdf, subtype_plot,
             width = 3, height = 3)

    } else {

      subtype_plot <- group_hist %>%
        filter(!is.na(mol_sub_group) & !grepl("CNS|EWS", mol_sub_group)) %>%
        dplyr::count(predicted_ancestry, mol_sub_group, name = "count", .drop = FALSE) %>%
        ggplot(aes(x = predicted_ancestry, y = mol_sub_group, fill = count)) +
        geom_tile(color = "black",
                  lwd = 1,
                  linetype = 1, show.legend = FALSE) +
        scale_fill_gradient(low="white", high="orangered") +
        geom_text(aes(label = count), color = "black", size = 4) +
        labs(x = NULL, y = NULL) +
        coord_fixed() +
        theme_minimal()
      ggsave(subtype_plot_pdf, subtype_plot,
             width = 3, height = 3)

    }
      
    kap_os <- survival_analysis(
    metadata  = group_hist[!grepl("To be classified", group_hist$mol_sub_group),],
    ind_var = var,
    test = "kap.meier",
    metadata_sample_col = "Kids_First_Biospecimen_ID",
    days_col = "OS_days",
    status_col = "OS_status"
    )
  
    readr::write_rds(kap_os, km_os_model_file)
    
    kap_efs <- survival_analysis(
    metadata  = group_hist[!grepl("To be classified", group_hist$mol_sub_group),],
    ind_var = var,
    test = "kap.meier",
    metadata_sample_col = "Kids_First_Biospecimen_ID",
    days_col = "EFS_days",
    status_col = "EFS_status"
    )
  
    readr::write_rds(kap_efs, km_efs_model_file)
    
    group_hist <- group_hist %>%
      mutate(predicted_ancestry = factor(predicted_ancestry,
                                            c("EUR", "AFR", "AMR", "EAS", "SAS"))) %>%
      mutate(race = factor(race,
                           c("White", "Asian", "Black/Afr American", "Other/Unknown"))) %>%
      mutate(ethnicity = fct_relevel(ethnicity,
                                c("Not Hispanic/Latino", "Hispanic/Latino")))
    
    if (group %in% c("Low-grade glioma")) {
      group_hist <- group_hist %>%
        dplyr::mutate(extent_of_tumor_resection = case_when(
          extent_of_tumor_resection %in% c('Not Reported', "Unavailable", "Not Applicable") ~ "Unavailable/Unreported",
          TRUE ~ extent_of_tumor_resection
        ))
    }
  
    add_model_os <- fit_save_model(group_hist[!grepl("To be classified", group_hist$mol_sub_group),],
                                    terms = ifelse(
                                        group %in% c("Low-grade glioma"),
                                        glue::glue("extent_of_tumor_resection+mol_sub_group+age_at_diagnosis_years+{var}"),
                                        ifelse(
                                            group %in% c("DIPG or DMG|Other high-grade glioma", "Ependymoma", "Medulloblastoma", "Mixed neuronal-glial tumor", "Atypical Teratoid Rhabdoid Tumor"), glue::glue("mol_sub_group+age_at_diagnosis_years+{var}"),
                                            glue::glue("age_at_diagnosis_years+{var}")
                                            )),
               coxph_os_model_file,
               "multivariate",
               years_col = "OS_years", status_col = "OS_status"
              )
    
    add_model_efs <- fit_save_model(group_hist[!grepl("To be classified", group_hist$mol_sub_group),],
                                    terms = ifelse(
                                        group %in% c("Low-grade glioma"),
                                        glue::glue("extent_of_tumor_resection+mol_sub_group+age_at_diagnosis_years+{var}"),
                                        ifelse(
                                            group %in% c("DIPG or DMG|Other high-grade glioma", "Ependymoma", "Medulloblastoma", "Mixed neuronal-glial tumor", "Atypical Teratoid Rhabdoid Tumor"), glue::glue("mol_sub_group+age_at_diagnosis_years+{var}"),
                                            glue::glue("age_at_diagnosis_years+{var}")
                                            )),
               coxph_efs_model_file,
               "multivariate",
               years_col = "EFS_years", status_col = "EFS_status"
              )
    
  }
}

```


```{r}

subtype_df <- data.frame(hist = c("Craniopharyngioma", "DIPG or DMG", "Other high-grade glioma",
                                  rep("Low-grade glioma", 3),
                                  "Mixed neuronal-glial tumor",
                                  rep("Medulloblastoma",3)),
                         names = c("cranio", "hgg", "hgg", "lgg", "lgg", 
                                   "lgg", "gng", "mb", "mb", "mb"),
                         subtype = c("ADAM", "K28", "WT", "WT",
                                     "BRAF", "Other", "Other",
                                     "Group4", "Group3", "SHH"))

variable <- c("predicted_ancestry", "race", "ethnicity", "EUR_status")

for (var in variable){
  
  for (i in 1:nrow(subtype_df)) {
    
    results_dir <- file.path(analysis_dir, "results", subtype_df$names[i])
    
    if (!dir.exists(results_dir)) {
      dir.create(results_dir)
      
      }
    
    km_os_model_file <- file.path(results_dir, 
                                      glue::glue("logrank_{subtype_df$names[i]}_{subtype_df$subtype[i]}_OS_{var}.RDS"))
  
    coxph_os_model_file <- ifelse(subtype_df$hist[i] %in% c("Low-grade glioma"),
                               file.path(results_dir,
                                        glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype[i]}_OS_additive_terms_subtype_resection_{var}.RDS")),
                               file.path(results_dir,
                                         glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype[i]}_OS_additive_terms_subtype_{var}.RDS"))
    )
  
      km_efs_model_file <- file.path(results_dir, 
                                      glue::glue("logrank_{subtype_df$names[i]}_{subtype_df$subtype[i]}_EFS_{var}.RDS"))
  
    coxph_efs_model_file <- ifelse(subtype_df$hist[i] %in% c("Low-grade glioma"),
                               file.path(results_dir,
                                        glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype[i]}_EFS_additive_terms_subtype_resection_{var}.RDS")),
                               file.path(results_dir,
                                         glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype[i]}_EFS_additive_terms_subtype_{var}.RDS"))
    )
  
      subtype_hist <- metadata %>%
      filter(plot_group == subtype_df$hist[i] & grepl(subtype_df$subtype[i], mol_sub_group))
      
      kap_os <- survival_analysis(
      metadata  = subtype_hist,
      ind_var = var,
      test = "kap.meier",
      metadata_sample_col = "Kids_First_Biospecimen_ID",
      days_col = "OS_days",
      status_col = "OS_status"
      )
      
      readr::write_rds(kap_os, km_os_model_file)
      
      kap_efs <- survival_analysis(
      metadata  = subtype_hist,
      ind_var = var,
      test = "kap.meier",
      metadata_sample_col = "Kids_First_Biospecimen_ID",
      days_col = "EFS_days",
      status_col = "EFS_status"
      )
      
      readr::write_rds(kap_efs, km_efs_model_file)
      
      if (subtype_df$hist[i] %in% c("Low-grade glioma")){
  
        subtype_hist <- subtype_hist %>%
          filter(!extent_of_tumor_resection %in% c('Not Reported', "Unavailable", "Not Applicable"))
      }
      
        subtype_hist <- subtype_hist %>%
            mutate(predicted_ancestry = factor(predicted_ancestry,
                                                  c("EUR", "AFR", "AMR", "EAS", "SAS"))) %>%
      mutate(race = factor(race,
                           c("White", "Asian", "Black/Afr American", "Other/Unknown"))) %>%
      mutate(ethnicity = factor(ethnicity,
                                c("Not Hispanic/Latino", "Hispanic/Latino")))
  
      add_model_os <- fit_save_model(subtype_hist,
                                    terms = ifelse(subtype_df$hist[i] %in% c("Low-grade glioma"),
                                                   glue::glue("extent_of_tumor_resection+age_at_diagnosis_years+{var}"),
                                                   glue::glue("age_at_diagnosis_years+{var}")),
                                     coxph_os_model_file,
                                     "multivariate",
                                     years_col = "OS_years", status_col = "OS_status"
                                    )
      
      add_model_efs <- fit_save_model(subtype_hist,
                                    terms = ifelse(subtype_df$hist[i] %in% c("Low-grade glioma"),
                                                   glue::glue("extent_of_tumor_resection+age_at_diagnosis_years+{var}"),
                                                   glue::glue("age_at_diagnosis_years+{var}")),
                                       coxph_efs_model_file,
                                       "multivariate",
                                       years_col = "EFS_years", status_col = "EFS_status"
                                      )
      
  }
    
}

```