---
title: 'Merge PBTA ancestry and histology data'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2023"
---
  
Load libraries and set directories
  
```{r load libraries and set directories}
library(data.table)
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "add-histologies")
results_dir <- file.path(analysis_dir, "results")
input_dir <- file.path(analysis_dir, "input")
```

Define input and output file paths

```{r set file paths}
# Somalier results
ancestry_file <- file.path(data_dir, "DEI_CBTN-PNOC_rerun_somalier_ancestry.tsv")

# files for converting IDs and filtering EU patients
pnoc_ids_file <- file.path(input_dir, "all_pnoc_normal_wgs.tsv")
hist_file <- file.path(data_dir, 'histologies.tsv')

## plot group designation file
plot_mapping_file <- file.path(input_dir, 'plot-mapping.tsv')

# hope project pathology review 
hope_file <- file.path(input_dir, "hope-non-hgg_path_review.tsv")

# output file
output_file <- file.path(results_dir, 'merged_ancestry_histology_data.tsv')
```

## Read in somalier results and histologies files

```{r load data}
ancestry <- read_tsv(ancestry_file)

hist <- read_tsv(hist_file, guess_max = 1000000)

```

## Add data columns from histology file

demographic data, pathology, and clinical data 
```{r add histology}
# match patient ID to BS ID in `res_all`
pnoc_ids <- read_tsv(pnoc_ids_file)

# Add participant IDs to `ancestry`
ancestry <- ancestry %>%
  left_join(hist[,c('Kids_First_Biospecimen_ID', 'Kids_First_Participant_ID', 'sample_type', "cohort_participant_id")], by = 'Kids_First_Biospecimen_ID') %>%
  dplyr::mutate(sample_type = case_when(
    is.na(sample_type) & Kids_First_Biospecimen_ID %in% pnoc_ids$Kids_First_Biospecimen_ID ~ 'Normal', 
    TRUE ~ sample_type)) %>%
  dplyr::filter(sample_type != 'Tumor')

# filter histologies file to exclude Normal samples
hist_tumor <- hist %>%
  filter(sample_type != 'Normal' & experimental_strategy == "WGS") %>%
  rename('Kids_First_Biospecimen_ID_tumor' = 'Kids_First_Biospecimen_ID')

# read in plot_mapping file to add plot group to results
plot_mapping <- read_tsv(plot_mapping_file)

# add histology data to `ancestry` by participant ID
ancestry <- ancestry %>%
  dplyr::left_join(hist_tumor[, c('Kids_First_Participant_ID', 'Kids_First_Biospecimen_ID_tumor', 'sample_id',
                              'reported_gender', 'race', 'ethnicity', 'broad_histology', 'cancer_group',
                              'molecular_subtype', 'tumor_descriptor', "CNS_region", 'extent_of_tumor_resection', 
                              'age_at_diagnosis_days', 'OS_days', 'OS_status', 'EFS_days', "EFS_event_type",
                     'age_at_chemo_start', 'age_at_radiation_start')], 
            by = 'Kids_First_Participant_ID') %>%
  dplyr::left_join(plot_mapping[,c('broad_histology', 'cancer_group', 'plot_group')], by = c('broad_histology', 'cancer_group'))
```


To select one normal-tumor pair per patient, select matched tumors in the following prioritization order: intial CNS tumor, recurrent, other. There are rare cases of different molecular subtypes being called from different tumors from the same patient. 

```{r}
initial_tumors <- ancestry %>%
  filter(tumor_descriptor == 'Initial CNS Tumor') %>%
  distinct(Kids_First_Participant_ID, .keep_all = T)

recurrent_tumors <- ancestry %>%
  filter(!Kids_First_Participant_ID %in% c(initial_tumors$Kids_First_Participant_ID) & tumor_descriptor == 'Recurrence') %>%
  distinct(Kids_First_Participant_ID, .keep_all = T)

other_tumors <- ancestry %>%
  filter(!Kids_First_Participant_ID %in% c(recurrent_tumors$Kids_First_Participant_ID, initial_tumors$Kids_First_Participant_ID)) %>%
  distinct(Kids_First_Participant_ID, .keep_all = T)
```

Merge data to obtain one row per patient 

```{r}
ancestry_unique <- initial_tumors %>%
  bind_rows(recurrent_tumors, other_tumors)
```


# Write merged data to file

```{r write output}
readr::write_tsv(ancestry_unique, file.path(results_dir, 'merged_ancestry_histology_data.tsv'))
```
