---
title: 'Merge PBTA ancestry and histology data'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2023"
---
  
Load libraries and set directories
  
```{r load libraries and set directories}
library(data.table)
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
setwd(root_dir)

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "add-histologies")
results_dir <- file.path(analysis_dir, "results")
input_dir <- file.path(analysis_dir, "input")
```

Define input and output file paths

```{r set file paths}
# Somalier results

ancestry_file <- file.path(input_dir, "DEI_CBTN-PNOC_rerun.somalier-ancestry.tsv")

# files for converting IDs and filtering EU patients
pbta_mapping <- file.path(input_dir, 'Broad_to_BS.txt')
pnoc_ids_file <- file.path(input_dir, "all_pnoc_normal_wgs.tsv")
eu_pt_file <- file.path(input_dir, 'Meyer_Tableau_Pull_11.8.2022.xlsm')
hist_file <- file.path(input_dir, 'histologies.tsv')

## plot group designation file
plot_mapping_file <- file.path(input_dir, 'plot-mapping.tsv')

# output file
output_file <- file.path(results_dir, 'merged_ancestry_histology_data.tsv')
```

## Read in somalier results

```{r read in results}
pbta_bs <- read_tsv(pbta_mapping)

ancestry <- read_tsv(ancestry_file) %>%
  rename('name' = '#sample_id') %>%  
  left_join(pbta_bs, by = 'name') %>%
  dplyr::mutate(Kids_First_Biospecimen_ID = case_when(
    grepl("BS", name) ~ name, 
    grepl("BS", `Kids First Biospecimen ID`) ~ `Kids First Biospecimen ID`,
    TRUE ~ NA
  )) %>%
  dplyr::filter(grepl('BS', Kids_First_Biospecimen_ID)) %>%
  dplyr::select(Kids_First_Biospecimen_ID, predicted_ancestry, given_ancestry,
                EAS_prob, AFR_prob, AMR_prob, SAS_prob, EUR_prob,
                PC1, PC2, PC3, PC4, PC5)
```
  
## Filter out EU patients
match patient IDs from EU list and histology file, filter out corresponding BS_ids from somalier results

```{r filter eu}
hist <- read_tsv(hist_file)

eu_pts <- readxl::read_excel(eu_pt_file)

eu_ids <- hist %>%
  filter(cohort_participant_id %in% eu_pts$`Participant`) %>%
  pull(Kids_First_Biospecimen_ID)

ancestry <- ancestry %>%
  filter(!Kids_First_Biospecimen_ID %in% eu_ids)
```


## Add data columns from histology file

demographic data, pathology, and clinical data 
```{r add histology}
# match patient ID to BS ID in `res_all`
pnoc_ids <- read_tsv(pnoc_ids_file)

ancestry <- ancestry %>%
  left_join(hist[,c('Kids_First_Biospecimen_ID', 'Kids_First_Participant_ID', 'sample_type', "cohort_participant_id")], by = 'Kids_First_Biospecimen_ID') %>%
  left_join(pnoc_ids[,c('Kids_First_Biospecimen_ID', 'Kids_First_Participant_ID')], by = 'Kids_First_Biospecimen_ID') %>%
  mutate(Kids_First_Participant_ID = case_when(
    !is.na(Kids_First_Participant_ID.x) ~ Kids_First_Participant_ID.x,
    !is.na(Kids_First_Participant_ID.y) ~ Kids_First_Participant_ID.y)
    ) %>%
  mutate(sample_type = ifelse(is.na(sample_type) & Kids_First_Biospecimen_ID %in% pnoc_ids$Kids_First_Biospecimen_ID, 'Normal', sample_type)) %>%
  filter(sample_type != 'Tumor') %>%
  dplyr::select(-c(Kids_First_Participant_ID.x, Kids_First_Participant_ID.y))

# filter histology file to exclude Normal samples
hist <- hist %>%
  filter(sample_type != 'Normal' & experimental_strategy == "WGS") %>%
  rename('Kids_First_Biospecimen_ID_tumor' = 'Kids_First_Biospecimen_ID')

plot_mapping <- read_tsv(plot_mapping_file)

# add histology data to `res_all` by participant ID
ancestry <- ancestry %>%
  left_join(hist[, c('Kids_First_Participant_ID', 'Kids_First_Biospecimen_ID_tumor',
                              'reported_gender', 'race', 'ethnicity', 'broad_histology', 'cancer_group',
                              'molecular_subtype', 'tumor_descriptor', "CNS_region", 'extent_of_tumor_resection', 
                              'age_at_diagnosis_days', 'OS_days', 'OS_status', 'EFS_days',
                     "age_at_chemo_start", "age_at_radiation_start")], 
            by = 'Kids_First_Participant_ID') %>%
  left_join(plot_mapping[,c('cancer_group', 'plot_group')], by = 'cancer_group')
```


To select one call per patient, select matched tumors in the following prioritization order: intial CNS tumor, recurrent, other. There are rare cases of different molecular subtypes being called from different tumors from the same patient. 

```{r}
initial_tumors <- ancestry %>%
  filter(tumor_descriptor == 'Initial CNS Tumor') %>%
  distinct(Kids_First_Participant_ID, .keep_all = T)

recurrent_tumors <- ancestry %>%
  filter(!Kids_First_Participant_ID %in% c(initial_tumors$Kids_First_Participant_ID) & tumor_descriptor == 'Recurrence') %>%
  distinct(Kids_First_Participant_ID, .keep_all = T)

other_tumors <- ancestry %>%
  filter(!Kids_First_Participant_ID %in% c(recurrent_tumors$Kids_First_Participant_ID, initial_tumors$Kids_First_Participant_ID)) %>%
  distinct(Kids_First_Participant_ID, .keep_all = T)
```

Merge data to obtain one row per patient 

```{r}
ancestry_unique <- initial_tumors %>%
  bind_rows(recurrent_tumors, other_tumors)
```

# Write merged data to file

```{r write output}
readr::write_tsv(ancestry_unique, file.path(results_dir, 'merged_ancestry_histology_data.tsv'))
```
