---
title: 'Merge ancestry and histology data'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2022-12-14"
---
  
Load libraries and set directories
  
```{r load libraries and set directories}
library(data.table)
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
setwd(root_dir)

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "add-histologies")
results_dir <- file.path(analysis_dir, "results")
input_dir <- file.path(analysis_dir, "input")
```

Define input and output file paths

```{r set file paths}
# Somalier results
cbtn_file <- file.path(input_dir, 'CBTN-WGS.somalier-ancestry.tsv')
pnoc_file <- file.path(input_dir, 'DEI_PNOC.somalier-ancestry.tsv')
pbta_file <- file.path(input_dir, 'DEI_PBTA.somalier-ancestry.tsv')
cbtn_run2_file <- file.path(input_dir, 'CBTN-PNOC_DEI.somalier-ancestry.tsv')

#f iles for converting IDs and filtering EU patients
pbta_mapping <- file.path(input_dir, 'Broad_to_BS.txt')
eu_pt_file <- file.path(input_dir, 'Meyer_Tableau_Pull_11.8.2022.xlsm')
hist_file <- file.path(input_dir, 'histologies.tsv')

# Methylation classification files
methyl_file <- file.path(input_dir, 'cbtn-dkfz-methylation-classification.csv')
subtype_file <- file.path(input_dir, 'methylation_to_OPC_subtype.csv')

# output file
output_file <- file.path(results_dir, 'merged_ancestry_histology_data.tsv')
```

## Read in somalier results

```{r read in results}
res_cbtn <- read_tsv(cbtn_file) %>%
  filter(grepl('BS', `#sample_id`)) %>%
  rename('Kids_First_Biospecimen_ID' = '#sample_id') %>%
  mutate(cohort = 'CBTN')

res_cbtn2 <- read_tsv(cbtn_run2_file) %>%
  filter(grepl('BS', `#sample_id`)) %>%
  rename('Kids_First_Biospecimen_ID' = '#sample_id') %>%
  mutate(cohort = 'CBTN')

res_pnoc <- read_tsv(pnoc_file) %>%
  filter(grepl('BS', `#sample_id`)) %>%
  rename('Kids_First_Biospecimen_ID' = '#sample_id') %>%
  mutate(cohort = 'PNOC')

# read in file to convert PBTA sample names to BS IDs
pbta_bs <- read_tsv(pbta_mapping)

res_pbta <- read_tsv(pbta_file) %>%
  rename('name' = '#sample_id') %>%
  left_join(pbta_bs, by = 'name') %>%
  rename('Kids_First_Biospecimen_ID' = 'Kids First Biospecimen ID') %>%
  select(-name) %>%
  relocate(`Kids_First_Biospecimen_ID`, .before = predicted_ancestry) %>%
  mutate(cohort = 'PBTA') %>%
  filter(grepl('BS', Kids_First_Biospecimen_ID))
```

Merge results across cohorts

```{r merge ancestry}
res_all <- res_cbtn %>%
  bind_rows(res_cbtn2, res_pnoc, res_pbta)
```
  
## Filter out EU patients
## match patient IDs from EU list and histology file, filter out corresponding BS_ids from somalier results
```{r filter eu}
hist <- read_tsv(hist_file)

eu_pts <- readxl::read_excel(eu_pt_file)

eu_ids <- hist %>%
  filter(cohort_participant_id %in% eu_pts$`Participant`) %>%
  pull(Kids_First_Biospecimen_ID)

res_all <- res_all %>%
  filter(!Kids_First_Biospecimen_ID %in% eu_ids)
```


## Add data columns from histology file
##
#demographic data, histology, cancer_group, molecular subtype, survival
```{r add histology}
# match patient ID to BS ID in `res_all`
res_all <- res_all %>%
  left_join(hist[,c('Kids_First_Biospecimen_ID', 'Kids_First_Participant_ID', 'sample_type')], by = 'Kids_First_Biospecimen_ID') %>%
  filter(sample_type != 'Tumor')

# filter histology file to exclude Normal samples and only retain one row per particpant ID
hist <- hist %>%
  filter(sample_type != 'Normal') %>%
  distinct(Kids_First_Participant_ID, .keep_all = T) %>%
  rename('Kids_First_Biospecimen_ID_tumor' = 'Kids_First_Biospecimen_ID')

# add histology data to `res_all` by participant ID
res_all <- res_all %>%
  left_join(hist[,c('Kids_First_Participant_ID', 'Kids_First_Biospecimen_ID_tumor', 'reported_gender', 
                    'race', 'broad_histology', 'cancer_group', 'molecular_subtype', 
                    'extent_of_tumor_resection', 'OS_days', 'OS_status')], 
            by = 'Kids_First_Participant_ID')
```

## Add methylation subclass and scores 
```{r add methylation}
#read in methylation file
methyl <- read_csv(methyl_file) %>%
  filter(sample_type == 'Tumor') %>%
  rename('Kids_First_Biospecimen_ID_tumor' = 'Kids_First_Biospecimen_ID')
head(methyl)

## add methylation subclass and score columns to `res_all`
res_all <- res_all %>%
  left_join(methyl[,c('Kids_First_Biospecimen_ID_tumor', 'dkfz_v12_methylation_subclass', 'dkfz_v12_methylation_subclass_score')], by = 'Kids_First_Biospecimen_ID_tumor') %>%
  filter(!is.na(Kids_First_Participant_ID)) %>%
  distinct(Kids_First_Participant_ID, .keep_all = T)
```

## Add column with corresponding OPC subtype for methylation subclasses, and create new aggregate subtype column prioritizing OPC subtype and taking methylation subtype when OPC not available
```{r}
subtypes <- read_csv(subtype_file) %>%
  rename('dkfz_v12_methylation_subclass' = 'Abbrevation')

#add OPC subtype corresponding to methylation subclasses
res_all <- res_all %>%
  left_join(subtypes[,c('dkfz_v12_methylation_subclass', 'OPC_molecular_subtype')], by = 'dkfz_v12_methylation_subclass') %>%
  rename('methylation_subtype' = 'OPC_molecular_subtype') %>%
  mutate(OPC_methyl_subtype = case_when(
    !is.na(molecular_subtype) & !grepl('To be classified', molecular_subtype) ~ molecular_subtype,
    !is.na(methylation_subtype) & dkfz_v12_methylation_subclass_score >= 0.8 ~ methylation_subtype,
    grepl('To be classified', molecular_subtype) ~ molecular_subtype,
    TRUE ~ NA_character_)
  )
```

# Write merged data to file

```{r write output}
readr::write_tsv(res_all, file.path(results_dir, 'merged_ancestry_histology_data.tsv'))
```
