---
title: 'Assess BRAF fusion breakpoints in PBTA ancestry cohort'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett, Jo Lynne Rokita
date: "2024"
---

This script calculates frequency of BRAF fusion breakpoint types by predicted ancestry in pLGG, BRAF fusion subtype patients in the PBTA ancestry cohort 
  
Load libraries and set directories
  
```{r load libraries and set directories}
library(data.table)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "braf-fusions")
results_dir <- file.path(analysis_dir, "results")
input_dir <- file.path(analysis_dir, "input")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(analysis_dir, "util", "heatmap_function.R"))
```

Set file paths

```{r}
fusion_file <- file.path(results_dir, "braf-fusions-exon-annotation.tsv")

fusion_dgd_file <- file.path(data_dir, "fusion-dgd.tsv.gz")

cohort_hist_file <- file.path(root_dir, "analyses", "survival", "results", "merged_ancestry_histology_survival.tsv")

hist_file <- file.path(data_dir, "histologies.tsv")
```

Load data

```{r}
braf_hist <- read_tsv(cohort_hist_file) %>%
  dplyr::filter(plot_group == "Low-grade glioma" & (grepl("-BRAF|wildtype|classified", molecular_subtype) | is.na(molecular_subtype))) %>%
  dplyr::mutate(predicted_ancestry = factor(predicted_ancestry))

hist <- read_tsv(hist_file)

# filter fusion file for in-frame fusions and duplications, and determine whether BRAF kinase domain is retained
braf_fusions <- read_tsv(fusion_file) %>%
  dplyr::filter(Fusion_Type == "in-frame",
         grepl("duplication", annots)) %>%
  mutate(keep = case_when(FusionName == "KIAA1549--BRAF" & DomainRetainedGene1B == "Yes" ~ "yes",
                          FusionName == "BRAF--KIAA1549" & DomainRetainedGene1A == "Yes" ~ "yes",
                          TRUE ~ "no")) %>%
  # The below samples have two different common breakpoints and one is low conf in arriba, so we will remove that one
  mutate(keep = case_when(Sample %in% c("BS_SDZY6RZ3", "BS_063ERW0R") & LeftBreakpoint_position == 138867975 ~ "no",
                            TRUE ~ keep))

braf_fusions_dgd <- read_tsv(fusion_dgd_file) %>%
  dplyr::filter(FusionName == "KIAA1549--BRAF" & !grepl("Tier 3|Tier 4", variant_tier)) %>%
  dplyr::rename(Kids_First_Biospecimen_ID = Sample) %>%
  dplyr::select(Kids_First_Biospecimen_ID, FusionName, `5_prime_region`, `3_prime_region`) %>%
  dplyr::mutate(`5_prime_region` = gsub("^Exon |^exon ", "", `5_prime_region`),
         `3_prime_region` = gsub("^Exon |^exon ", "", `3_prime_region`),                        
         breakpoint_exons = paste(`5_prime_region`, `3_prime_region`, sep = ":")) %>%
    # re-format exon numbers to match those in pbta fusion file
  dplyr::mutate(breakpoint_exons = str_replace(
                                              str_replace(breakpoint_exons,
                                               ":8", ":08"),
                                              ":9", ":09"),
         breakpoint_type = case_when(breakpoint_exons %in% c("16:09", "15:09", "16:11", "18:10") ~ "common",
                                     TRUE ~ "rare/novel"),
         keep = "yes") %>%
  left_join(hist[,c("Kids_First_Participant_ID", "Kids_First_Biospecimen_ID")])
```

Consolidate all common and rare exon breakpoints per patient

```{r}
common_fusions_pbta <- braf_fusions %>%
  dplyr::filter(breakpoint_exons %in% c("16:09", "15:09", "16:11", "18:10") & keep == "yes") %>%
  distinct(Kids_First_Participant_ID, breakpoint_exons, .keep_all = T) %>%
  group_by(Kids_First_Participant_ID) %>%
  summarise(breakpoint_exons_common = str_c(breakpoint_exons, collapse = ";"))

common_fusions <- braf_fusions_dgd %>%
  dplyr::filter(breakpoint_exons %in% c("16:09", "15:09", "16:11", "18:10") & keep == "yes") %>%
  distinct(Kids_First_Participant_ID, breakpoint_exons, .keep_all = T) %>%
  group_by(Kids_First_Participant_ID) %>%
  summarise(breakpoint_exons_common = str_c(breakpoint_exons, collapse = ";")) %>%
  bind_rows(common_fusions_pbta) %>%
  distinct(Kids_First_Participant_ID, breakpoint_exons_common, .keep_all = T) %>%
  group_by(Kids_First_Participant_ID) %>%
  summarise(breakpoint_exons_common = str_c(breakpoint_exons_common, collapse = ";"))

rare_novel_fusions_pbta <- braf_fusions %>%
  dplyr::filter(!breakpoint_exons %in% c("16:09", "15:09", "16:11", "18:10") & keep == "yes") %>%
  distinct(Kids_First_Participant_ID, breakpoint_exons, .keep_all = T) %>%
  group_by(Kids_First_Participant_ID) %>%
  summarise(breakpoint_exons_rare = str_c(breakpoint_exons, collapse = ";"))

rare_novel_fusions <- braf_fusions_dgd %>%
  dplyr::filter(!breakpoint_exons %in% c("16:09", "15:09", "16:11", "18:10") & keep == "yes") %>%
  distinct(Kids_First_Participant_ID, breakpoint_exons, .keep_all = T) %>%
  group_by(Kids_First_Participant_ID) %>%
  summarise(breakpoint_exons_rare = str_c(breakpoint_exons, collapse = ";")) %>%
  bind_rows(rare_novel_fusions_pbta) %>%
  distinct(Kids_First_Participant_ID, breakpoint_exons_rare, .keep_all = T) %>%
  group_by(Kids_First_Participant_ID) %>%
  summarise(breakpoint_exons_rare = str_c(breakpoint_exons_rare, collapse = ";"))

fusions_all <- common_fusions %>%
  bind_rows(rare_novel_fusions) %>%
  dplyr::mutate(breakpoint_type = case_when(
    !is.na(breakpoint_exons_common) & !is.na(breakpoint_exons_rare) ~ NA_character_,
    !is.na(breakpoint_exons_rare) ~ "rare",
    !is.na(breakpoint_exons_common) ~ "common",
    TRUE ~ NA_character_
  )) %>%
  # Assign breakpoint group (unique common breakpoint or broad rare group)
  dplyr::mutate(breakpoint_group = case_when(
    !is.na(breakpoint_exons_common) & !is.na(breakpoint_exons_rare) ~ NA_character_,
    breakpoint_exons_common == "16:09" ~ "16:09",
    breakpoint_exons_common == "15:09" ~ "15:09",
    breakpoint_exons_common == "16:11" ~ "16:11",
    breakpoint_exons_common == "18:10" ~ "18:10",
    !is.na(breakpoint_exons_rare) ~ "rare",
    TRUE ~ NA_character_
  )) %>%
  write_tsv(file.path(results_dir, "braf-fusion-breakpoints-by-patient.tsv"))
```

Merge breakpoint data to braf hist

```{r}
# add high conf methylation subtypes to the braf hist
scores <- hist %>%
  filter(dkfz_v12_methylation_subclass_score >= 0.8,
         # grab only PA subtypes
         grepl("PA_", dkfz_v12_methylation_subclass)) %>%
  dplyr::select("match_id", "dkfz_v12_methylation_subclass") %>%
  group_by(match_id) %>%
  summarise(dkfz_v12_methylation_subclass = str_c(unique(dkfz_v12_methylation_subclass), collapse = ";")) %>%
  unique()

braf_hist <- braf_hist %>%
  dplyr::rename(Kids_First_Biospecimen_ID_normal = Kids_First_Biospecimen_ID,
                Kids_First_Biospecimen_ID = Kids_First_Biospecimen_ID_tumor) %>%
  left_join(fusions_all) %>%
  dplyr::filter(!is.na(breakpoint_type)) %>%
  left_join(hist[,c("Kids_First_Biospecimen_ID", "match_id")]) %>%
  left_join(scores)
```

How many patients do we have breakpoint data for? 

```{r}
nrow(braf_hist)
```

```{r}
round(table(braf_hist$predicted_ancestry, braf_hist$breakpoint_type)/as.vector(table(braf_hist$predicted_ancestry)) * 100, 2)
```

## Common breakpoint summary

What is the distribution of common breakpoints by ancestry? 

```{r}
table(braf_hist$breakpoint_exons_common, braf_hist$predicted_ancestry)
```

Calculate fusion breakpoint frequency by predicted ancestry and Fisher's exact test p-value 

```{r}
common_breakpoints <- c("16:09", "15:09", "16:11", "18:10")

common_df <- braf_hist %>%
  dplyr::filter(breakpoint_exons_common %in% common_breakpoints) %>%
  group_by(breakpoint_exons_common, predicted_ancestry) %>%
  summarise(count = n()) %>%
  group_by(predicted_ancestry) %>%
  mutate(total_count = sum(count),
         percentage = round(count / total_count * 100, 1)) %>%
  dplyr::mutate(ct_perc = glue::glue("{count} ({percentage}%)")) %>%
  dplyr::select(breakpoint_exons_common, predicted_ancestry, ct_perc) %>%
  spread(predicted_ancestry, ct_perc, fill = 0)

common_df <- common_df %>%
  mutate(
    p = map_dbl(breakpoint_exons_common, ~round(fisher.test(table(grepl(.x, braf_hist$breakpoint_exons_common), braf_hist$predicted_ancestry))$p.value, 4))
  )

write_tsv(common_df, 
          file.path(results_dir, "lgg-braf-fusion-common-breakpoint-freq.tsv"))
```

Calculate breakpoint group odds ratios and p-values across predicted ancestries, and plot

```{r}
group_ht <- plot_enr(braf_hist, 
                     "breakpoint_group", "predicted_ancestry",
                     var2_names = c("AFR", "AMR", "EAS", "EUR", "SAS"),
                     var1_names = c("15:09", "16:09", "16:11", "18:10", "rare"),
                     padjust = TRUE)

# plot enrichment results
pdf(file.path(plot_dir, "breakpoint_group_ancestry_ct_enr_heatmap.pdf"),
    height = 3, width = 4)

draw(group_ht)

invisible(dev.off())

draw(group_ht)
```

Print rare/novel fusion breakpoint count by predicted ancestry

```{r}
table(braf_hist$breakpoint_exons_rare, braf_hist$predicted_ancestry)
```

Create heatmap of rare breakpoints by ancestry
```{r}
pdf(file.path(plot_dir, "rare_breakpoints_by_ancestry.pdf"),
    height = 3, width = 4)

rare_ct <- table(braf_hist$breakpoint_exons_rare[!is.na(braf_hist$breakpoint_exons_rare)],
                 braf_hist$predicted_ancestry[!is.na(braf_hist$breakpoint_exons_rare)])

col_fun = colorRamp2(c(0, ceiling(max(rare_ct))), c("white", "orangered"))

rare_ht <- Heatmap(rare_ct,
            name = "Count",
            cluster_rows = F,
            cluster_columns = F,
            rect_gp = gpar(col = "black", lwd = 2),
            col = col_fun,
            cell_fun = function(j, i, x, y, width, height, fill) {
              grid.text(sprintf("%s", rare_ct[i, j]), x, y, gp = gpar(fontsize = 12))
            })

rare_ht

invisible(dev.off())

rare_ht
```

Print ancestry by methyl subtype

```{r}
table(braf_hist$dkfz_v12_methylation_subclass, braf_hist$predicted_ancestry)
```


Print breakpoint by methyl subtype

```{r}
table(braf_hist$breakpoint_type, braf_hist$dkfz_v12_methylation_subclass)
fisher.test(table(braf_hist$breakpoint_type, braf_hist$dkfz_v12_methylation_subclass), simulate.p.value = T)
```

Print breakpoint by methyl subtype

```{r}
table(braf_hist$breakpoint_group, braf_hist$dkfz_v12_methylation_subclass)
fisher.test(table(braf_hist$breakpoint_group, braf_hist$dkfz_v12_methylation_subclass), simulate.p.value = T)
```

Plot methyl subtype by breakpoint group

```{r}
methyl_ht <- plot_enr(braf_hist[!is.na(braf_hist$dkfz_v12_methylation_subclass),], 
                     "breakpoint_group", "dkfz_v12_methylation_subclass",
                     var1_names = c("15:09", "16:09", "16:11", "18:10", "rare"),
                     var2_names = sort(unique(braf_hist$dkfz_v12_methylation_subclass)),
                     padjust = TRUE)

# plot enrichment results
pdf(file.path(plot_dir, "breakpoint_group_methyl_subtype_ct_enr_heatmap.pdf"),
    height = 3, width = 4)

draw(methyl_ht)

invisible(dev.off())

draw(methyl_ht)
```

Assess distribution of resection level in patients by breakpoint type (common vs. rare)

```{r}
round(table(braf_hist$breakpoint_type, braf_hist$extent_of_tumor_resection)/as.vector(table(braf_hist$breakpoint_type)), 2)

fisher.test(table(braf_hist$breakpoint_type, braf_hist$extent_of_tumor_resection))
```

Assess distribution of resection level in patients by breakpoint group

```{r}
round(table(braf_hist$breakpoint_group, braf_hist$extent_of_tumor_resection)/as.vector(table(braf_hist$breakpoint_group)), 2)

fisher.test(table(braf_hist$breakpoint_group, braf_hist$extent_of_tumor_resection), simulate.p.value = T)
```

Plot extent of tumor resection by breakpoint group

```{r}
braf_hist <- braf_hist %>%
  dplyr::mutate(resection = str_replace(extent_of_tumor_resection, " resection", ""))

resection_ht <- plot_enr(braf_hist[!grepl("Not Reported", braf_hist$resection),], 
                     "breakpoint_group", "resection",
                     var1_names = c("15:09", "16:09", "16:11", "18:10", "rare"),
                     var2_names = c("Gross/Near total", "Partial", "Biopsy only"),
                     padjust = TRUE)

# plot enrichment results
pdf(file.path(plot_dir, "breakpoint_group_resection_ct_enr_heatmap.pdf"),
    height = 3, width = 4)

draw(resection_ht)

invisible(dev.off())

draw(resection_ht)
```

Assess distribution of tumor CNS region in patients by breakpoint type and group

```{r}
round(table(braf_hist$breakpoint_type, braf_hist$CNS_region)/as.vector(table(braf_hist$breakpoint_type)), 3) * 100
fisher.test(table(braf_hist$breakpoint_type, braf_hist$CNS_region), simulate.p.value = T)
```


```{r}
round(table(braf_hist$breakpoint_group, braf_hist$CNS_region)/as.vector(table(braf_hist$breakpoint_group)), 3) * 100

fisher.test(table(braf_hist$breakpoint_group, braf_hist$CNS_region), simulate.p.value = T)
```
Plot CNS region by breakpoint group

```{r}
region_ht <- plot_enr(braf_hist[braf_hist$CNS_region != "Mixed",], 
                     "breakpoint_group", "CNS_region",
                     var1_names = c("15:09", "16:09", "16:11", "18:10", "rare"),
                     var2_names = sort(unique(braf_hist$CNS_region[braf_hist$CNS_region != "Mixed"])),
                     padjust = TRUE)

# plot enrichment results
pdf(file.path(plot_dir, "breakpoint_group_region_ct_enr_heatmap.pdf"),
    height = 3, width = 6.5)

draw(region_ht)

invisible(dev.off())

draw(region_ht)
```

Confirm that CNS region and degree of tumor resection are significantly associated: 

```{r}
round(table(braf_hist$CNS_region, braf_hist$extent_of_tumor_resection)/as.vector(table(braf_hist$CNS_region)), 3) * 100

fisher.test(table(braf_hist$CNS_region, braf_hist$extent_of_tumor_resection), simulate.p.value= T)
```

```{r}
hist <- read_tsv(file.path(data_dir, "histologies.tsv"))

diagnoses <- read_tsv(file.path(input_dir, "lgg_braf_fusion_final_diagnosis.txt"))

braf_hist <- braf_hist %>%
  left_join(hist %>% dplyr::select(Kids_First_Biospecimen_ID, pathology_free_text_diagnosis)) %>%
  dplyr::mutate(diagnosis = case_when(
    grepl("pilocytic/pilomyxoid", pathology_free_text_diagnosis) ~ "pilocytic/pilomyxoid astrocytoma",
    grepl("pilocytic|pilocystic|pilocytioc", pathology_free_text_diagnosis) | Kids_First_Biospecimen_ID %in% diagnoses$Kids_First_Biospecimen_ID[diagnoses$final_diagnosis == "pilocytic astrocytoma"] ~ "pilocytic astrocytoma",
    grepl("pilomyxoid", pathology_free_text_diagnosis) | Kids_First_Biospecimen_ID %in% diagnoses$Kids_First_Biospecimen_ID[diagnoses$final_diagnosis == "pilomyxoid astrocytoma"] ~ "pilomyxoid astrocytoma",
    grepl("fibrillary", pathology_free_text_diagnosis) | Kids_First_Biospecimen_ID %in% diagnoses$Kids_First_Biospecimen_ID[diagnoses$final_diagnosis == "fibrillary astrocytoma"] ~ "fibrillary astrocytoma",
    TRUE ~ NA_character_
  ))

diagnosis_ht = plot_enr(braf_hist, var1 = "diagnosis", var2 = "breakpoint_group",
                         var1_names = c("fibrillary astrocytoma", 
                                        "pilocytic astrocytoma", 
                                        "pilocytic/pilomyxoid astrocytoma",
                                        "pilomyxoid astrocytoma"), 
                        var2_names = c("15:09", "16:09", "16:11", "18:10", "rare"),
                         padjust = TRUE)

pdf(file.path(plot_dir, "lgg_braf_fusion_breakpoint_group_diagnosis_enr_heatmap.pdf"),
    width = 6, height = 3)

draw(diagnosis_ht)

dev.off()
```

Save braf-specific hist file with breakpoint annotation 

```{r}
braf_hist %>%
  dplyr::select(-resection) %>%
  write_tsv(file.path(results_dir, "lgg-braf-fusion-breakpoint-annotation.tsv"))
```

Print session info

```{r}
sessionInfo()
```