---
title: 'Assess BRAF fusion breakpoints in PBTA ancestry cohort'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Jo Lynne Rokita
date: "2024"
---

This script calculates frequency of BRAF fusion breakpoint types by predicted ancestry in pLGG, BRAF fusion subtype patients in the PBTA ancestry cohort 
  
Load libraries and set directories
  
```{r load libraries and set directories}
library(data.table)
library(tidyverse)
library(readxl)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "braf-fusions")
results_dir <- file.path(analysis_dir, "results")
input_dir <- file.path(analysis_dir, "input")
plot_dir <- file.path(analysis_dir, "plots")

```

Set file paths

```{r}
fusion_file <- file.path(results_dir, "braf-fusion-breakpoints-by-patient.tsv")
hist_file <- file.path(data_dir, "histologies.tsv")
cluster_file <- file.path(input_dir, "20231003_merged_imaging_clustering_risk_stratification.xlsx")
fusion_dgd_file <- file.path(data_dir, "fusion-dgd.tsv.gz")
```

Load subtype and imaging cluster data

```{r load imaging data and append proper events and subtypes}
# histology file - remove normals and derived cell lines
hist <- read_tsv(hist_file, guess_max = 100000) %>%
  filter(!is.na(pathology_diagnosis),
         composition != "Derived Cell Line")

# fusion file
break_df <- read_tsv(fusion_file) %>%
  left_join(unique(hist[,c("cohort_participant_id", "Kids_First_Participant_ID")])) %>%
  select(-Kids_First_Participant_ID)

# imaging cluster data
image_clusters <- readxl::read_excel(cluster_file) %>%
    dplyr::rename(cohort_participant_id = SubjectID,
                age_at_diagnosis_days_original = age_at_event_days.1_y,
                event_id_original = cbtn_specimen_group_id_x,
                age_at_imaging_days = Days_at_Imaging_y) %>%
  # update cluster assignments
  dplyr::mutate(`Cluster Assignment` = case_when(`Cluster Assignment` == 0 ~ 1,
                                                 `Cluster Assignment` == 1 ~ 2,
                                                 `Cluster Assignment` == 2 ~ 3))

# choose earliest age for the sample subtype to utilize since imaging is taken closes to diagnosis
earliest_age_subtypes <- image_clusters %>%
  select(cohort_participant_id, age_at_imaging_days, age_at_diagnosis_days_original, event_id_original) %>%
  left_join(unique(hist[,c("cohort_participant_id", "sample_id", "molecular_subtype", "age_at_diagnosis_days", "age_at_event_days")]), by = c("cohort_participant_id")) %>%
  group_by(cohort_participant_id) %>%
  # select earliest event
  summarise(age_at_event_days = min(age_at_event_days)) %>%
  left_join(unique(hist[,c("cohort_participant_id", "sample_id", "molecular_subtype", "age_at_diagnosis_days", "age_at_event_days")]), by = c("cohort_participant_id", "age_at_event_days")) %>%
  dplyr::mutate(molecular_subtype = gsub("^LGG, |^GNG, |^SEGA, ", "", molecular_subtype)) %>%  group_by(cohort_participant_id) %>%
  # collapse events, update molecular subtypes
  summarise(age_at_diagnosis_collapse = str_c(unique(age_at_diagnosis_days), collapse = ";"),
            earliest_events = str_c(unique(sample_id), collapse = ";"),
            molecular_subtype = str_c(unique(molecular_subtype), collapse = ";")) %>%
  mutate(molecular_subtype = case_when(grepl("HGG|;To be|, wildtype|To be classified;|wildtype;", molecular_subtype) ~ gsub(";To be classified|; wildtype|To be classified;|wildtype; |HGG, To be classified;", "", molecular_subtype), TRUE ~ molecular_subtype)) %>%
  full_join(image_clusters) %>%
  dplyr::mutate(age_dx_match = ifelse(age_at_diagnosis_days_original == age_at_diagnosis_collapse, "yes", "no")) %>%
  rowwise() %>%
  dplyr::mutate(event_id_match = if_else(sapply(strsplit(earliest_events, ";"), function(x) event_id_original %in% x), "yes", "no"))
```

Merge breakpoint data to hist

```{r}
imaging_plus_fusions <- earliest_age_subtypes %>%
  left_join(break_df)
```

Add breakpoint_type column (common vs rare) and breakpoint_group column (common breakpoint type and rare merged)

```{r}
imaging_plus_fusions <- imaging_plus_fusions %>%
  # if NA subtype and there is a breakpoint, add KIAA1549--BRAF:
  dplyr::mutate(molecular_subtype = case_when(is.na(molecular_subtype) & !is.na(breakpoint_type) ~ "KIAA1549-BRAF",
                                              TRUE ~ molecular_subtype)) %>%
  # update col order
  select(cohort_participant_id, age_at_imaging_days, age_at_diagnosis_days_original, age_at_diagnosis_collapse, age_dx_match, event_id_original, earliest_events, event_id_match, Cohort, `Distance to 0`, `Distance to 1`, `Distance to 2`, `Cluster Assignment`, molecular_subtype, starts_with("breakpoint")) %>%
  write_tsv(file.path(results_dir, "imaging_clusters_plus_subtypes_breakpoints.tsv"))
```


How many patients do we have breakpoint data for? 

```{r}
imaging_plus_fusions %>%
  filter(!is.na(breakpoint_type)) %>%
  nrow()
```

## Subtypes type by cluster

What is the distribution of type of molecular subtypes by cluster? 

```{r}
table(imaging_plus_fusions$molecular_subtype, imaging_plus_fusions$`Cluster Assignment`)
```


## Breakpoint type by cluster

What is the distribution of type of breakpoints by cluster? 

```{r}
table(imaging_plus_fusions$breakpoint_type, imaging_plus_fusions$`Cluster Assignment`)
```

```{r stats on common vs rare}
type_df <- imaging_plus_fusions %>%
  filter(!is.na(breakpoint_type)) %>%
  group_by(breakpoint_type, `Cluster Assignment`) %>%
  summarise(count = n()) %>%
  group_by(breakpoint_type) %>%
  mutate(total_count = sum(count),
         percentage = round(count / total_count * 100, 1)) %>%
  dplyr::mutate(ct_perc = glue::glue("{count} ({percentage}%)")) %>%
  dplyr::select(breakpoint_type, `Cluster Assignment`, ct_perc) %>%
  spread(`Cluster Assignment`, ct_perc, fill = 0)

type_df <- type_df %>%
  mutate(
    p = map_dbl(breakpoint_type, ~round(fisher.test(table(grepl(.x, imaging_plus_fusions$breakpoint_type), imaging_plus_fusions$`Cluster Assignment`))$p.value, 10))
  )
type_df

write_tsv(type_df, 
          file.path(results_dir, "lgg-braf-fusion-common-breakpoint-type-image-clusters.tsv"))

```


Assess distribution of breakpoint type by imaging cluster (common vs. rare)

```{r}
fisher.test(table(imaging_plus_fusions$breakpoint_type, imaging_plus_fusions$`Cluster Assignment`))
```

## Common breakpoint summary

What is the distribution of common breakpoints by cluster? 

```{r}
table(imaging_plus_fusions$breakpoint_exons_common, imaging_plus_fusions$`Cluster Assignment`)
```

Calculate fusion breakpoint frequency by predicted ancestry and Fisher's exact test p-value 

```{r}
common_df <- imaging_plus_fusions %>%
  dplyr::filter(!is.na(breakpoint_exons_common)) %>%
  group_by(breakpoint_exons_common, `Cluster Assignment`) %>%
  summarise(count = n()) %>%
  group_by(breakpoint_exons_common) %>%
  mutate(total_count = sum(count),
         percentage = round(count / total_count * 100, 1)) %>%
  dplyr::mutate(ct_perc = glue::glue("{count} ({percentage}%)")) %>%
  dplyr::select(breakpoint_exons_common, `Cluster Assignment`, ct_perc) %>%
  spread(`Cluster Assignment`, ct_perc, fill = 0)

common_df <- common_df %>%
  mutate(
    p = map_dbl(breakpoint_exons_common, ~round(fisher.test(table(grepl(.x, imaging_plus_fusions$breakpoint_exons_common), imaging_plus_fusions$`Cluster Assignment`))$p.value, 10))
  )

common_df

write_tsv(common_df, 
          file.path(results_dir, "lgg-braf-fusion-common-breakpoint-freq-image-clusters.tsv"))

```

Print rare/novel fusion breakpoint count by predicted ancestry

```{r}
table(imaging_plus_fusions$breakpoint_exons_rare, imaging_plus_fusions$`Cluster Assignment`)

```

Print session info

```{r}
sessionInfo()

```